Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-27 manuscript/manuscript.tex
changed site choice for decay figure

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing the nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM's)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation~\citep{si2008empirical}\jdbcomment{check that this should be Si Quang and not Quang} are discussed in the last section of the Results.

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter\jdbcomment{cite Haddox et al paper}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
We will denote such models as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

\begin{figure}
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as measured by \citet{doud2016accurate}.
The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
For the GY94 model, the graphs are identical for all sites since this models does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
\jdbcomment{Can you choose something instead of site 348 where the preferences and asymptotic divergence distributions look similar? Here they look very different since there are 6 Leu codons and 1 Met codon}
}
\end{figure}

>>>>

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94+$\Gamma\omega$ model with the commonly used F3X4 stationary state has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing the nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9,500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM's)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data.
The number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian \citep{lartillot2004bayesian, rodrigue2014site} or maximum-likelihood estimation~\citep{si2008empirical} are discussed in the last section of the Results.

Importantly, these two strategies for modeling purifying selection are not mutually exclusive.
Mutation-selection models such as ExpCM can still incorporate an $\omega$ parameter~\citep{haddox2017mapping}, which now represents the relative rate of nonsynonymous to synonymous substitution \emph{after} accounting for the constraints due to the site-specific stationary state.
This $\omega$ parameter for an ExpCM can be drawn from a distribution just like for GY94-style models. 
We will denote such models as ExpCM+$\Gamma\omega$.
\ref{fig:model_feature}C shows the full spectrum of models that incorporate all combinations of gamma-distributed $\omega$ and site-specific stationary states.

@@

\begin{figure}
\centerline{\includegraphics[width=0.90\textwidth]{figures/decay.pdf}}
\caption{\label{fig:decay}
\textbf{Effect of stationary state and $\Gamma\omega$ rate variation on predicted asymptotic sequence divergence.}
The logo plots at top show the amino-acid preferences for some sites in an H1 influenza hemagglutinin protein as measured by \citet{doud2016accurate}.
The graphs show the expected amino-acid identity at that site for two sequences separated by a branch of the indicated length (\ref{eq:f}).
For the GY94 model, the graphs are identical for all sites since this models does not have site-specific parameters; the same is true for GY94+$\Gamma\omega$.
The graphs do differ among sites if we use the amino-acid preferences to calculate a different $\omega_r$ value for each site $r$ in a GY94 framework~\citep[\ref{eq:w_r};][]{spielman2015relationship}.
However, all GY94 models, including the one with site-specific $\omega_r$ values, approach the same asymptote since they all have the same stationary state.
But the ExpCM has different asymptotes for different sites since it accounts for how amino-acid preferences lead to site-specific stationary states.
\skhcomment{Currently, this plot is made using the hybridDoud preferences. I should change the numbering to reflect WSN sequential numbering}
}
\end{figure}
